---
title: "Abalone"
output: html_document
---
Example use of Logistic Regression (glm) function in R

Read in the data. We have to add the column names by hand.

```{r}
library('ggplot2')
d <- read.table('abalone.data.txt.gz',header=FALSE,sep=',',
                stringsAsFactors=TRUE)
# add the column names
colnames(d) <- c('Sex', 'Length', 'Diameter', 'Height', 'WholeWeight',
   'ShuckedWeight', 'VisceraWeight', 'ShellWeight', 'Rings')
dim(d) 
summary(d) 
```

Define the outcome variable to be whether or not the abalone has ten rings or more.
```{r}
# define 'old' as more than 10 rings (more than ~11.5 years old)
d$old <- d$Rings>=10
table(d$old)/nrow(d)
```

Look at the Sex variable
```{r}
summary(d$Sex)
```

Check whether Sex separates the data
```{r}
tab = table(truth=d$Rings, sex=d$Sex)
rs = rowSums(tab)
tab/rowSums(tab)
```

Do the test/train split
```{r}
set.seed(2352) # set random seed for reproducibility

# do the random split (25% held out for test), put the label back into the data frame
d$isTest <- runif(nrow(d))<0.25
d$dataLabel <- ifelse(d$isTest,"test data","train data")
```

Try a non-invasive prediction procedure first. 
(Only the measurements we can take without dissecting the abalone)
```{r}
nonInvasiveVars <- c('Sex', 'Length', 'Diameter', 'Height', 'WholeWeight')
# the model formula
fNonInvasive <- paste('old',paste(nonInvasiveVars,collapse=' + '),sep=' ~ ')
fNonInvasive
```

Run the model. 
```{r}
model1 <- glm(fNonInvasive,data=d[!d$isTest,],family=binomial(link='logit'))
summary(model1)  
```

Look at the model's predictions on the data.
```{r}
d$model1 <- predict(model1,newdata=d,type='response')
head(d[,c("old","model1")])
```

Calculate pseudo-R-Squared on training data
```{r}
# training pseudo-R2
1 - model1$deviance/model1$null.deviance
```

Plot the predictions for the training data. 
```{r}
dtrain = d[!d$isTest,]  # makes the following calls easier
ggplot(dtrain, aes(x=model1, color=old)) + geom_density()
```

Create the confusion matrix
```{r}
# confusion matrix

cmat = table(truth=dtrain$old, pred=dtrain$model1 > 0.5)
cmat
```

Calculate the accuracy (training data)
```{r}
accuracy = (cmat[1,1] + cmat[2,2])/sum(cmat)
accuracy
```

Convenience functions to report performance
```{r}
deviance <- function(truth,pred,epsilon=0) {
  pred = pmax(pred, epsilon)
  pred = pmin(pred, 1-epsilon)
  S = 0.0 # assumed log-likelihood of saturated model
  -2*(sum(ifelse(truth,log(pred),log(1-pred)))-S)
}

deviance(dtrain$old, dtrain$model1)
model1$deviance
```

```{r}

reportStats <- function(d,test,modelName,title) {
  dSub <- d[d$isTest==test,,drop=FALSE]
  tab <- table(truth=dSub$old,pred=dSub[,modelName]>0.5)
  accuracy <- (tab[1,1] + tab[2,2])/sum(tab)
  note = ifelse(test,'test','train')
  print(paste('   ',note,'accuracy',modelName,format(accuracy,digits=2)))
  residual.deviance <- deviance(dSub[,'old'],dSub[,modelName])
  null.deviance <- deviance(dSub[,'old'],mean(dSub[,'old']))
  print(paste("   model explained a",
              format((1-residual.deviance/null.deviance),digits=2),
            "fraction of the variation on",note))  
}

report <- function(d,modelName,title) {
  print("***********")
  print(paste("model",modelName,title))
  reportStats(d,FALSE,modelName,title)
  reportStats(d,TRUE,modelName,title)
  print(ggplot(data=d,aes_string(x=modelName,color='old')) + 
    geom_density() + facet_wrap(~dataLabel,ncol=1,scales='free_y') +
    ggtitle(title))
  print("***********")
}
```

Performance report for both the training and test data
```{r}
# report the performance of the model on hold-out
report(d, "model1", "Logistic Regression: Non-invasive Model")
```

Set up a model that includes invasive measurments
```{r}
invasiveVars <- c('ShuckedWeight', 'VisceraWeight', 'ShellWeight')
fAllVars <- paste('old',paste(c(nonInvasiveVars,invasiveVars),collapse=' + '),sep=' ~ ')
fAllVars
```

Fit the model
```{r}
model2 <- glm(fAllVars,data=d[!d$isTest,],family=binomial(link='logit'))
d$model2 <- predict(model2,newdata=d,type='response')
summary(model2)
```

Report performance statistics
```{r}
report(d,'model2', "invasive model logistic regression")
```

