---
title: "DplyrDependencies"
author: "Win-Vector LLC"
date: "11/30/2017"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
#bibliography: skeleton.bib
link-citations: yes
---

In [an earlier note](https://github.com/WinVector/Examples/blob/master/dplyr/Dependencies.md) we exhibited a non-signalling result corruption in `dplyr` `0.7.4`. In this note we
demonstrate the [`seplyr`](https://winvector.github.io/seplyr/) work-around.

Re-establish up our example:

```{r setup}
packageVersion("dplyr")

my_db <- DBI::dbConnect(RSQLite::SQLite(),
                        ":memory:")
d <- dplyr::copy_to(
  my_db, 
  data.frame(
    valuesA = c("A", NA, NA),
    valuesB = c("B", NA, NA),
    canUseFix1 = c(TRUE, TRUE, FALSE),
    fix1 = c('Fix_1_V1', "Fix_1_V2", "Fix_1_V3"),
    canUseFix2 = c(FALSE, FALSE, TRUE),
    fix2 = c('Fix_2_V1', "Fix_2_V2", "Fix_2_V3"),
    stringsAsFactors = FALSE),
  'd', 
  temporary = TRUE, overwrite = TRUE)
knitr::kable(dplyr::collect(d))
```

[`seplyr`](https://winvector.github.io/seplyr/) has a fix/work-around for the earlier issue: automatically break up the steps into safe blocks ([announcement](http://www.win-vector.com/blog/2017/11/win-vector-llc-announces-new-big-data-in-r-tools/); here we are using the development [`seplyr`](https://winvector.github.io/seplyr/) `0.5.1` version of [`mutate_se()`](https://winvector.github.io/seplyr/reference/mutate_se.html)).

```{r goodresult2}
library("seplyr")
packageVersion("seplyr")

d %.>% 
  mutate_nse(
    ., 
    valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                      fix1, valuesA),
    valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                      fix2, valuesA),
    valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                      fix1, valuesB),
    valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                      fix2, valuesB),
    mutate_nse_printPlan = TRUE) %.>% 
  select_se(., c("valuesA", "valuesB")) %.>% 
  dplyr::collect(.) %.>% 
  knitr::kable(.)
```

We now have a correct result (all cells filled).

`seplyr` used safe statement re-ordering to break the calculation into the minimum
number of blocks/groups that have no in-block dependencies between statements (note
this is more efficient that merely introducing a new mutate each first time a new
value is used).

We can slow that down and see how the underlying planning
functions break the assignments down into a small
number of safe blocks (here we are using the development [`wrapr`](https://winvector.github.io/wrapr/) `1.0.2` function
[`qae()`](https://winvector.github.io/wrapr/reference/qae.html)).

```{r goodresult2s}
packageVersion("wrapr")

steps <- qae(
  valuesA := ifelse(is.na(valuesA) & canUseFix1, 
                    fix1, valuesA),
  valuesA := ifelse(is.na(valuesA) & canUseFix2, 
                    fix2, valuesA),
  valuesB := ifelse(is.na(valuesB) & canUseFix1, 
                    fix1, valuesB),
  valuesB := ifelse(is.na(valuesB) & canUseFix2, 
                    fix2, valuesB))
print(steps)
plan <- partition_mutate_se(steps)
print(plan)

d %.>% 
  mutate_seb(., plan) %.>% 
  select_se(., c("valuesA", "valuesB")) %.>% 
  dplyr::collect(.) %.>% 
  knitr::kable(.)
```

Note that the current `CRAN` versions of [`wrapr`](https://CRAN.R-project.org/package=wrapr) and [`seplyr`](https://CRAN.R-project.org/package=seplyr) *already*
implement the above work-around.  Just some of the conveniences such
as `mutate_nse_printPlan = TRUE` and `qae()` require the development versions
of these packages.

