---
title: "Partitioning Mutate, Example 3"
author: "John Mount, Win-Vector LLC"
date: "2017-12-28"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
#bibliography: skeleton.bib
link-citations: yes
---

```{r setupa, include=FALSE}
library("tufte")
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

This third article shows our second example being processed by the [`rquery`](https://github.com/WinVector/rquery) big data scale relational data operator system (currently in development).

```{r sed, include=FALSE}
library("wrapr")
library("rquery")

sc <- 
  sparklyr::spark_connect(version = '2.2.0', 
                          master = "local")
dL <- data.frame(rowNum = 1:5,
                 a_1 = "",
                 a_2 = "",
                 b_1 = "",
                 b_2 = "",
                 c_1 = "",
                 c_2 = "",
                 d_1 = "",
                 d_2 = "",
                 e_1 = "",
                 e_2 = "",
                 stringsAsFactors = FALSE)
d <- rquery::dbi_copy_to(sc, dL, 
                         table_name = 'd',
                         overwrite = FALSE,
                         temporary = TRUE)

```

We will repeat the steps from [Partitioning Mutate, Example 2](http://winvector.github.io/FluidData/partition_mutate_ex2.html), using only 
the `rquery` package and `DBI`/`sparklyr` (no `dplyr`).

```{r p1}
library("wrapr")
library("rquery")

class(sc)
class(d)

d %.>% 
  to_sql(., sc) %.>%
  DBI::dbGetQuery(sc, .) %.>%
  knitr::kable(.)
```


```{r dL, comment = ""}
dQ <- d %.>%
  extend_se(.,
            if_else_block(
              testexpr =
                "rand()>=0.5",
              thenexprs = qae(
                a_1 := 'treatment',
                a_2 := 'control'),
              elseexprs =  qae(
                a_1 := 'control',
                a_2 := 'treatment'))) %.>%
  select_columns(., c("rowNum", "a_1", "a_2"))

cat(format(dQ))
```

Notice the `rquery` `extend_se` command accepts the `if_else_block` *and*
partitions it into conflict-free segments.  Also the `rquery` presentation lets
the user inspect the operation plan before attempting execution.



```{r dLex}
sql <- to_sql(dQ, sc)
DBI::dbGetQuery(sc, sql) %.>%
  knitr::kable(.)
```

The underlying `SQL` is fairly involved, but can be performant at big-data scale.

```{r dS, comment=""}
cat(sql)
```

Links
-----

[Win-Vector LLC](http://www.win-vector.com/) supplies a number of open-source
[`R`](https://www.r-project.org) packages for working effectively with big data.
These include:

  * **[wrapr](https://winvector.github.io/wrapr/)**: supplies code re-writing tools that make coding *over* ["non standard evaluation"](http://adv-r.had.co.nz/Computing-on-the-language.html) interfaces (such as `dplyr`) *much* easier.
  * **[cdata](https://winvector.github.io/cdata/)**: supplies pivot/un-pivot functionality at big data scale.
  * **[rquery](https://github.com/WinVector/rquery)**: (in development) big data scale relational data operators.
  * **[seplyr](https://winvector.github.io/seplyr/)**: supplies improved interfaces for many data manipulation tasks.
  * **[replyr](https://winvector.github.io/replyr/)**: supplies tools and patches for using `dplyr` on big data.

Partitioning mutate articles:

  * **[Partitioning Mutate](http://winvector.github.io/FluidData/partition_mutate.html)**: basic example.
  * **[Partitioning Mutate, Example 2](http://winvector.github.io/FluidData/partition_mutate_ex2.html)**: `ifelse` example.
  * **[Partitioning Mutate, Example 3](http://winvector.github.io/FluidData/partition_mutate_ex3.html)** [`rquery`](https://github.com/WinVector/rquery) example.

Topics such as the above are often discussed on the [Win-Vector blog](http://www.win-vector.com/blog/).

```{r cleanup, include=FALSE}
sparklyr::spark_disconnect(sc)
```
