---
title: "Partitioning Mutate, Example 2"
author: "John Mount, Win-Vector LLC"
date: "2017-11-24"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
#bibliography: skeleton.bib
link-citations: yes
---

```{r setupa, include=FALSE}
library("tufte")
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

`Sparklyr` with its `dplyr` translations allows `R` to perform the heavy lifting that has traditionally been the exclusive domain of proprietary systems like `SAS`.
In general, `dplyr` is good at handling intermediate variables in the mutate function so users don't need to think about it.
However, some of that breaks down when the processing is done on the `Apache Spark` side. [Win-Vector LLC](http://win-vector.com)
developed the [`seplyr`](https://winvector.github.io/seplyr/) package to use with clients to mitigate some of these situations.^[And we distribute the package as open-source to give back to the `R` community.]
In this article we will demonstrate we `seplyr` functions: `if_else_device()` and ``partition_mutate_qt()`.^[Note: these functions are new code and under development in version `0.5.0` 
of [`seplyr`](https://github.com/WinVector/seplyr) (and not yet available 
in the [CRAN release version](https://CRAN.R-project.org/package=seplyr)).]


This is a follow-on example building on our ["Partitioning Mutate" article](http://winvector.github.io/FluidData/partition_mutate.html).
showing a larger block sequence based on swaps.^[The source code for this article can be found [here](https://github.com/WinVector/WinVector.github.io/blob/master/FluidData/partition_mutate_ex2.Rmd).]  For motivation and context please see
[the first article](http://winvector.github.io/FluidData/partition_mutate.html).

Please consider the following example data (on a remote `Spark` cluster).

```{r sed, include=FALSE}
library("seplyr")

sc <- 
  sparklyr::spark_connect(version = '2.2.0', 
                          master = "local")
dL <- data.frame(rowNum = 1:5,
                 a_1 = "",
                 a_2 = "",
                 b_1 = "",
                 b_2 = "",
                 c_1 = "",
                 c_2 = "",
                 d_1 = "",
                 d_2 = "",
                 e_1 = "",
                 e_2 = "",
                 stringsAsFactors = FALSE)
d <- dplyr::copy_to(sc, dL, 
                    name = 'd',
                    overwrite = TRUE,
                    temporary = TRUE)

```

```{r p1}
class(d)
d %.>% 
  dplyr::collect(.) %.>%  # https://github.com/tidyverse/dplyr/issues/3216
  knitr::kable(.)
```

It is often necessary to simulate block commands with `ifelse()`
style functionality.  

For our example: suppose we wish to assign columns in a complementary
to treatment and control design^[Abraham Wald designed some sequential analysis procedures
in this way as Nina Zumel [remarked](https://github.com/WinVector/ODSCWest2017/tree/master/MythsOfDataScience). Another string example is conditionals where you are trying to vary on a per-row basis which column is assigned to, instead of varying what value is assigned from.]
And further suppose we want to keep the random variables driving our decisions
around for diagnosis and debugging.

To write such a procedure in pure `dplyr` we might simulate block with code such
as the following^[Only showing work on the `a` group right now. We are assuming
we want to perform this task on all the grouped letter columns.]

```{r dL}
library("seplyr")

plan <- if_else_device(
  "rand()>=0.5",
  thenexprs = c(
    "a_1" := "'treatment'",
    "a_2" := "'control'"),
  elseexprs =  c(
    "a_1" := "'control'",
    "a_2" := "'treatment'"))  %.>% 
  partition_mutate_se(.)


d %.>% 
  mutate_seb(., plan) %.>% 
  select_se(., grepdf('^ifebtest_.*', ., invert=TRUE)) %.>% 
  dplyr::collect(.) %.>% 
  knitr::kable(.)
```

Above we are using the indent notation to indicate the code-blocks we are simulating
with row-wise `if(){}else{}` blocks.^[For more on this concept, please see: [https://winvector.github.io/seplyr/reference/if_else_device.html](https://winvector.github.io/seplyr/reference/if_else_device.html).]

Now our goal was to perform this same operation on each of the 5 letter groups.

We do this easily as follows:^[A better overall design would be to use 
[`cdata::moveValuesToRowsN()`](https://winvector.github.io/cdata/reference/moveValuesToRowsN.html),
then perform a single bulk operation on rows, and then pivot/transpose back
with [`cdata::moveValuesToColumnsN()`](https://winvector.github.io/cdata/reference/moveValuesToColumnsN.html).
But let's see how we simply work with a problem at hand.]

```{r dLong}
plan <- lapply(c('a', 'b', 'c', 'd', 'e'),
               function(gi) {
                 if_else_device(
                   "rand()>=0.5",
                   thenexprs = c(
                     paste0(gi, "_1") := "'treatment'",
                     paste0(gi, "_2") := "'control'"),
                   elseexprs =  c(
                     paste0(gi, "_1") := "'control'",
                     paste0(gi, "_2") := "'treatment'"))
               }) %.>% 
  unlist(.) %.>% 
  partition_mutate_se(.)

d %.>% 
  mutate_seb(., plan) %.>% 
  select_se(., grepdf('^ifebtest_.*', ., invert=TRUE)) %.>% 
  dplyr::collect(.) %.>% 
  knitr::kable(.)
```

Please keep in mind: we are using a very simple and regular sequence only
for purposes of illustration.  The intent is to show the types of issues
one runs into when standing-up non-trivial applications in `Sparklyr`.

The purpose of [`seplyr::partition_mutate_qt()`](https://github.com/WinVector/seplyr) 
is to re-arrange statements and break them into blocks of non-dependent statements (no
statement in a block depends on any other in the same block, and all value dependencies
are respected by the block order).  `seplyr::partition_mutate_qt()` if further defined to do this
in a performant manner.^[That is to pick a small number of blocks, in our case the plan consisted of
`r length(plan)` blocks.  The simple method of introducing a block boundary at each first use
of derived value (without statement re-ordering) would create a very much larger set of blocks 
(which cause problems of their own).  In particular the impression code and comments
of [upcoming `dplyr` fix](https://github.com/tidyverse/dbplyr/commit/36a44cd4b5f70bc06fb004e7787157165766d091) appear
to indicate an undesirable large number of blocks solution.]

Without such partition planning the current version of `dplyr` (`r packageVersion("dplyr")`)
the results of `dplyr::mutate()` do not seem to be well-defined when values are created
and re-used in the same `dplyr::mutate()` block.  This is not a currently documented limitation, but
it is present:

```{r bad}
ex <- dplyr::mutate(d,
                    condition_tmp = rand()>=0.5,
                    a_1 = ifelse( condition_tmp, 
                                  'treatment', 
                                  a_1),
                    a_2 = ifelse( condition_tmp, 
                                  'control', 
                                  a_2),
                    a_1 = ifelse( !( condition_tmp ), 
                                  'control', 
                                  a_1),
                    a_2 = ifelse( !( condition_tmp ), 
                                  'treatment', 
                                  a_2))

knitr::kable(dplyr::collect(dplyr::select(ex, a_1, a_2)))
```

Notice above the many `NA` columns, which are errors.^[Note: no mere re-ordering of the statements would give this result.]

```{r sq}
dplyr::show_query(ex)
```

Looking at the query we see that one of the conditional statements is missing (notice only 3 case statements, not 4):^[Likely 
the `dplyr` `SQL` generator does not perform a correct live-value analysis and therefor gets fooled into thinking a statement can
safely be eliminated (when it can not). `seplyr::partition_mutate_qt()` performs a correct live value calculation and make sure
`dplyr::mutate()` is only seeing trivial blocks (blocks where no value depends on any calculation in the same block).]

Conclusion
----------

`seplyr::if_else_device()` and `seplyr::partition_mutate_qt()` type capability is
essential for executing non-trivial code at scale in `Sparklyr`.

[Win-Vector LLC](http://www.win-vector.com/) supplies a number of open-source
[`R`](https://www.r-project.org) packages for working effectively with big data.
These include:

  * **[wrapr](https://winvector.github.io/wrapr/)**: supplies code re-writing tools that make coding *over* `dplyr` much easier.
  * **[cdata](https://winvector.github.io/cdata/)**: supplies pivot/un-pivot functionality at big data scale.
  * **[seplyr](https://winvector.github.io/seplyr/)**: supplies improved interfaces for many data manipulation tasks.
  * **[replyr](https://winvector.github.io/replyr/)**: supplies tools and patches for using `dplyr` on big data.

Topics such as the above are often discussed on the [Win-Vector blog](http://www.win-vector.com/blog/).

```{r cleanup, include=FALSE}
sparklyr::spark_disconnect(sc)
```
