---
title: "isolet"
output: html_document
---


```{r}
library('ggplot2')
library('randomForest')
```

First we load the data. The test and training sets have already been given in this
set, but I'll put them together with a test/train marker as we usually have
```{r}
# test and training sets already given
dTrain = read.table("isolet1+2+3+4.data.gz",
              header=FALSE,sep=',',
              stringsAsFactors=FALSE,blank.lines.skip=TRUE)
dTrain$isTest <- FALSE
dTrain$testlabel="train"

dTest = read.table("isolet5.data.gz",
              header=FALSE,sep=',',
              stringsAsFactors=FALSE,blank.lines.skip=TRUE)
dTest$isTest = TRUE
dTest$testlabel="test"
d = rbind(dTrain,dTest) 
# make "train" the first level, so it shows up first in the plots
rank = ifelse(d$isTest, 2, 1)
d$testlabel = reorder(d$testlabel, rank, FUN=mean)
rm(list=c('dTest','dTrain')) # clean up the environment a little by deleting the smaller sets

dim(d)
colnames(d)[1:5]
```

Make the frame a little more user-friendly
```{r}
# column 618 designates the letter as a number
unique(d$V618)
letters # a predefined (by R) vector
```

```{r}
# so let's map the numbers to letters
d$V618 = letters[d$V618]
summary(as.factor(d$V618))
# rename the column
colnames(d)[618] = "letter"
```


Define variables and output
```{r}
vars = colnames(d)[1:617]
yColumn = 'isN'
```

We are just going to work on the letters 'm' and 'n'.
'n' is our target class
```{r}
# Subset to just the utterances 'n', and 'm'
d = d[d$letter %in% c('m','n'),,drop=FALSE]
# 'n' is our target class
d[,yColumn] = d[,'letter']=='n'
table(d[,yColumn])/nrow(d)
```

First, some convenience functions for performance stats
```{r}
deviance = function(truth,pred,epsilon=0) {
  pred = pmax(pred, epsilon)
  pred = pmin(pred, 1-epsilon)
  S = 0.0 # assumed log-likelihood of saturated model
  -2*(sum(ifelse(truth,log(pred),log(1-pred)))-S)
}

pseudo_Rsquared = function(truth, pred, epsilon) {
  dev = deviance(truth, pred, epsilon)
  null.dev = deviance(truth, mean(pred), epsilon)
  1 - (dev/null.dev)
}

accuracy = function(truth, pred) {
  # confusion matrix
  cmat = table(truth, pred>0.5)
  sum(diag(cmat))/sum(cmat)
}

reportStats = function(d,test,modelName,title,epsilon=1e-02) {
  dSub = d[d$isTest==test,,drop=FALSE]
  acc = accuracy(dSub[,yColumn], dSub[,modelName])
  r2 = pseudo_Rsquared(dSub[,yColumn], dSub[,modelName], epsilon)
  note = ifelse(test,'test','train')
  print(paste('\t',note,'accuracy',modelName,format(acc,digits=2)))
  print(paste("\tmodel explained a",
              format(r2,digits=2),
            "fraction of the variation on",note))  
}

report <- function(d,modelName,title,epsilon=1.0e-2) {
  print("***********")
  reportStats(d, FALSE, modelName, title, epsilon)
  reportStats(d, TRUE, modelName, title, epsilon)
  print("***********")
}
```

Run the random forest model
```{r}
set.seed(23424) # set seed for reproducibility 
                # RF algorithm is randomized
# it's better not to use formulas for random forest. Use x and y
model1 = randomForest(x=d[!d$isTest,vars],
                      y=as.factor(d[!d$isTest,yColumn]),
                      importance=TRUE)
model1
```

Do the predictions 
```{r}
d$model1 = predict(model1,newdata=d,type='prob')[,'TRUE',drop=TRUE]
```

Report performance
```{r}
# double density plot
ggplot(d, aes_string(x="model1", color=yColumn)) + geom_density() + facet_wrap(~testlabel, ncol=1)

report(d,'model1',"randomForest")
```

We can examine the variable importances
```{r}
# first plot it (top 30 or so)
varImpPlot(model1, type=1) # just look at mean decrease accuracy
```

Get the variable importances
```{r}
# get the variable importances
vImps = importance(model1, class='TRUE')
head(vImps)
```

Sort the variable importances
```{r}
ord = order(vImps[,3], decreasing=TRUE) # sort, descending order
vord = vImps[ord, 3]
head(vord)
```

Get the top 30
```{r}
# limit to the top 30
topvars = names(vord[1:30])
```

Make a reduced model
```{r}
model2 = randomForest(x=d[!d$isTest,topvars],y=as.factor(d[!d$isTest,yColumn]))
d$model2 = predict(model2,newdata=d,type='prob')[,'TRUE',drop=TRUE]                      
report(d,'model2',"randomForest")
```

(Interestingly, did better on deviance)

