---
title: "Plotting Keras Performance Trajectories"
author: "John Mount"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
---

Below is the standard performance trajectory plot from `Keras`.

```{r setup}
library("ggplot2")
library("WVPlots")
library("keras")

h <- readRDS("historyobject.rds")
plot(h)
```

This is our `WVPlots` plot.


```{r wvplots}
d <- readRDS("metricsframe.rds")

WVPlots::plot_Keras_fit_trajectory(
  d, 
  title = "model performance by epoch, dataset, and measure")
```

Obviously this plot needs some training to interpret, but that is pretty much the case for all visualizations.

The ideas of this plot include:

  * All facets have the same interpretation: "up is better." This makes it easier to relate the shapes in one pane to the other.
  * The only horizontal solid curve rendered is the validation (or hold-out) performance.  The validation performance is the primary item of interest, so it gets a unique presentation.
  * The training performance is represented as the upper boundary of the shaded region and with the pluses and vertical strokes.
  * A vertical line is added at the best epoch picked by the "minus binary cross entropy" metric (this can be turned off or changed).
  * An optional additional training over-fit penalized dashed curve is added.  The default is: this is the validation score minus `10%` of the excess generalization error (the difference in training and validation performance).


