---
title: "Accum.Rmd"
author: "John Mount"
date: "July 24, 2015"
output: html_document
---


Function to populate a row of data.

```{r}
mkRow <- function(nCol) {
  x <- as.list(rnorm(nCol))
  x[[1]] <- ifelse(x[[1]]>0,'pos','neg')
  names(x) <- paste('x',seq_len(nCol),sep='.')
  x
}
```

The common wrong-way to accumulate the rows of data into a single data frame.

```{r}
mkFrameForLoop <- function(nRow,nCol) {
  d <- c()
  for(i in seq_len(nRow)) {
    ri <- mkRow(nCol)
    di <- data.frame(ri,
                     stringsAsFactors=FALSE)
    d <- rbind(d,di)
  }
  d
}
```

Timing showing the quadratic runtime.

```{r}
library('microbenchmark')
library('ggplot2')
set.seed(23525) # make run more repeatable

nCol <- 10
timeSeq <- seq.int(100,2000,100)
timings <- list()
for(nRow in timeSeq) {
  ti <- microbenchmark(
    mkFrameForLoop(nRow,nCol),
    times=10)
  ti <- data.frame(ti,
                   stringsAsFactors=FALSE)
  ti$nRow <- nRow
  ti$nCol <- nCol
  timings[[length(timings)+1]] <- ti
}
timings <- do.call(rbind,timings)

ggplot(data=timings,aes(x=nRow,y=time)) +
  geom_point() + geom_smooth()

timings1 <- timings
```

A few roughly equivilent right ways to accumlate the rows.

```{r}
# Simplest fix, collect the data in a list and
# grow the list.  Exploits the fact that R can mutate
# common objects when object visibility is limited.
mkFrameForList <- function(nRow,nCol,verbose=FALSE) {
  d <- as.list(seq_len(nRow))
  for(i in seq_len(nRow)) {
    ri <- mkRow(nCol)
    di <- data.frame(ri,
                     stringsAsFactors=FALSE)
    d[[i]] <- di
    if(verbose) {
      print(pryr::address(d))
    }
  }
  do.call(rbind,d)
}

# Cleanest fix- wrap procedure in a function and
# use lapply.
mkFrameList <- function(nRow,nCol) {
  d <- lapply(seq_len(nRow),function(i) {
    ri <- mkRow(nCol)
    data.frame(ri,
                     stringsAsFactors=FALSE)
  })
  do.call(rbind,d)
}

# Mucking with environments fix.  Environments
# are mutable and tend to be faster than lists.
# However any "improvement" is surely drown out
# by the costs of conversion to list and rbind,
# so the notational compromise is not worth it.
mkFrameEnv <- function(nRow,nCol) {
  e <- new.env(hash=TRUE,parent=emptyenv())
  for(i in seq_len(nRow)) {
    ri <- mkRow(nCol)
    di <- data.frame(ri,
                     stringsAsFactors=FALSE)
    assign(as.character(i),di,envir=e)
  }
  do.call(rbind,as.list(e))
}

# Another possibility, working in place.
# In a mutable language this pre-allocation
# would work.  In R 3.2.0 the fact that data frames
# are collections of columns (not rows) means the
# "limitted visibility" allows true in-place 
# alteration of objects hack is not available and
# this code is slow.
mkFrameInPlace <- function(nRow,nCol) {
  r1 <- mkRow(nCol)
  d <- data.frame(lapply(r1,
                         function(col) {
                           replicate(nRow,col)
                           }),
                  stringsAsFactors=FALSE)
  if(nRow>1) {
    for(i in seq.int(2,nRow,1)) {
      ri <- mkRow(nCol)
      di <- data.frame(ri,
                     stringsAsFactors=FALSE)
      d[i,] <- di
    }
  }
  d
}
```

Confirm value getting altered in place (effiency depends on interior columns also not chaning address, which is also the case).

```{r}
mkFrameForList(10,5,TRUE)
```

Get more timings and plots.

```{r}
timings <- list()
for(nRow in timeSeq) {
  ti <- microbenchmark(
    mkFrameForList(nRow,nCol),
    mkFrameList(nRow,nCol),
    times=10)
  ti <- data.frame(ti,
                   stringsAsFactors=FALSE)
  ti$nRow <- nRow
  ti$nCol <- nCol
  timings[[length(timings)+1]] <- ti
}
timings <- do.call(rbind,timings)

ggplot(data=timings,aes(x=nRow,y=time,color=expr)) +
  geom_point() + geom_smooth() 

timings <- rbind(timings,timings1)

ggplot(data=timings,aes(x=nRow,y=time,color=expr)) +
  geom_point() + geom_smooth() 
```

Show the in-place alteration of objects in a simpler setting.

```{r}
computeSquares <- function(n,
                           messUpVisibility,
                           usePRYR=FALSE) {
  # pre-allocate v
  # (doesn't actually help!)
  v <- 1:n
  if(messUpVisibility) {
     vLast <- v
  }
  # print details of v
  .Internal(inspect(v))
  if(usePRYR) {
    print(pryr::address(v))
  }
  for(i in 1:n) {
    v[[i]] <- i^2
    if(messUpVisibility) {
      vLast <- v
    }
    # print details of v
    .Internal(inspect(v))
    if(usePRYR) {
      print(pryr::address(v))
    }
  }
  v
}
```

Show how if the value associated with v is visible only to the variable name "v" that R will start performing in-place replacement (making calculation much cheaper).

```{r}
computeSquares(5,FALSE)
```

Show how if the value associated with v is visible to more than one variable that R will not performing in-place replacement (making calcultion much more expensive).

```{r}
computeSquares(5,TRUE)
```

Check if we can use PRYR for addresses in this case.

```{r}
computeSquares(5,FALSE,usePRYR=TRUE)
```

