---
title: "APASig"
author: "Win-Vector LLC"
date: "October 2, 2016"
output: html_document
---

Examples for [Proofing statistics in papers](http://www.win-vector.com/blog/2016/10/proofing-statistics-in-papers/).  Uses formatting [R](https://cran.r-project.org) package [sigr](https://github.com/WinVector/sigr). Please see [here](https://web2.uconn.edu/writingcenter/pdf/Reporting_Statistics.pdf) and [here](http://my.ilstu.edu/~jhkahn/apastats.html) for some notes.

```{r fns}
library('sigr') # devtools::install_github('WinVector/sigr')
```

Example showing we can get the items reported by `summary(model)` into one
well-behaved string.

```{r demo}
library('ggplot2')
d <- data.frame(x=0.2*(1:20))
d$y <- cos(d$x)
model <- lm(y~x,data=d)
d$prediction <- predict(model,newdata=d)

ggplot(data=d,aes(x=prediction,y=y)) +
  geom_point() + geom_abline() 
print(summary(model))
```

Examples showing how to get the summary from the model.

```{r results='asis'}
formatFTestFromModel(model,pSmallCutoff=1.0e-12)
```

Or frome the data.

```{r results='asis'}
formatFTestFromData(predictions=d$prediction,y=d$y,
                    pSmallCutoff=1.0e-12)
```

Separate extracted example from https://web2.uconn.edu/writingcenter/pdf/Reporting_Statistics.pdf.

```{r results='asis'}
formatFTest(numdf=2,dendf=55,FValue=5.56)
```

Looks like statCheck checks the p-value, but not the R-squared.

```{r}
library('statcheck')
s <- statcheck('(R2=.38, F(2,55)=5.56, p < .01)')
```
```{r results='asis', comment=NA}
print(xtable::xtable(t(s)),type='html')
```

Check ours.

```{r}
s <- statcheck('(R2=0.17, F(2,55)=5.56, p=0.00632)')
```
```{r results='asis', comment=NA}
print(xtable::xtable(t(s)),type='html')
```

